---
title: Creating continuous coloured maps with osmplotr
author: Thierry Onkelinx
date: '2017-07-17'
slug: creating-maps-with-osmplotr
categories: ["reproducible-research", "maps"]
tags: ["maps", "osm", "openstreetmap"]
banner: post/2017-07-17-creating-maps-with-osmplotr_files/figure-html/osmplotr-1.png
params:
  cellsize: 20
  scale: 2
---

During [useR!2017](https://user2017.brussels/) I attended a talk by Mark Padgham titled [Maps are data, so why no plot data on a map](https://channel9.msdn.com/Events/useR-international-R-User-conferences/useR-International-R-User-2017-Conference/Maps-are-data-so-why-plot-data-on-a-map). In this blog post I will recreate the map from an earlier blogpost on [estimating densities from a point pattern](../../06/estimating-densities-from-a-point-pattern).

Let's start by loading all required packages.

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  cache = TRUE,
  autodep = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r load-packages}
library(sp)
library(MASS)
library(dplyr)
library(ggmap)
library(osmplotr)
```

I rearranged the code from the previous blog post in two chunks. The first chunk calculate the density and is required for both the `ggmap` and the `osmplotr` solution. See the [earlier blog post](../../06/estimating-densities-from-a-point-pattern) for the details on what the code does.

```{r create-object}
points <- read.delim("../../data/20170628/points.txt", sep = " ")
crs_wgs84 <- CRS("+init=epsg:4326")
crs_lambert <- CRS("+init=epsg:31370")
coordinates(points) <- ~lon + lat
points@proj4string <- crs_wgs84
points_lambert <- spTransform(points, crs_lambert)
xlim <- range(pretty(points_lambert$lon)) + c(-100, 100)
ylim <- range(pretty(points_lambert$lat)) + c(-100, 100)
n <- c(
  diff(xlim),
  diff(ylim)
) / params$cellsize + 1
dens <- kde2d(
  x = points_lambert$lon,
  y = points_lambert$lat,
  n = n,
  lims = c(xlim, ylim)
)
dens <- expand.grid(
  lon = dens$x,
  lat = dens$y
) %>%
  mutate(
    lon = mean(lon) + params$scale * (lon - mean(lon)),
    lat = mean(lat) + params$scale * (lat - mean(lat))
  ) %>%
  mutate(
    density = as.vector(dens$z) * length(points_lambert),
    id = seq_along(density)
  )
```

The second chunk converts the density to a `SpatialPolygonsDataFrame` so we can plot it with `ggmap`.

```{r ggmap-dens}
dx <- diff(dens$lon[1:2])
dy <- diff(unique(dens$lat)[1:2])
dens_gg <- dens %>%
  merge(
    data.frame(
      x = dx * (c(0, 0, 1, 1, 0) - 0.5),
      y = dy * (c(0, 1, 1, 0, 0) - 0.5)
    )
  ) %>%
  mutate(
    lon = lon + x,
    lat = lat + y
  )
coordinates(dens_gg) <- ~lon + lat
dens_gg@proj4string <- crs_lambert
dens_wgs <- spTransform(dens_gg, crs_wgs84) %>%
  as.data.frame()
summary(dens_wgs)
```

The next chunk gets the satelite image and overlays is with the density map (figure \@ref(fig:ggmap)). The density layer needs to be somewhat transparant in order to see the background image.

```{r ggmap, fig.cap = "`ggmap` image of density"}
ggmap(
  get_map(
    location = c(lon = mean(points$lon), lat = mean(points$lat)),
    zoom = 17,
    maptype = "satellite",
    source = "google"
  )
) +
  geom_polygon(
    data = dens_wgs, 
    aes(group = id, fill = density), 
    alpha = 0.5
  ) +
  scale_fill_gradientn(
    colours = rev(rainbow(100, start = 0, end = .7)),
    limits = c(0, NA)
  )
```

The `osmplotr` package uses a set of points to create the colours. `x` and `y` hold the longitude an latitude, `z` the variable for the colour value.

```{r osm-dens}
coordinates(dens) <- ~lon + lat
dens@proj4string <- crs_lambert
dens_oms <- spTransform(dens, crs_wgs84)
dataset <- as.data.frame(dens_oms) %>%
  transmute(x = lon, y = lat, z = density)
```

The next step is to download the required [OpenStreetMap (OSM)](https://www.openstreetmap.org) data. Here I selected all roads and paths ("highway" is OSM terminology) and all the buildings. 

```{r extraxt-osm}
bb <- bbox(dens_oms)
roads <- extract_osm_objects(key = 'highway', bbox = bb)
buildings <- extract_osm_objects(key = 'building', bbox = bb)
```

Finaly, the map is created. `osm_basemap()` defines the extent and the background colour. `add_osm_surface()` adds the coloured layers of roads and buildings. I used `adjust_colours(-0.2)` to make the colours of roads a bit darker.

```{r osmplotr, fig.cap = "`osmplotr` image of density"}
osm_basemap(bbox = bb, bg = "gray95") %>%
  add_osm_surface(
    buildings,
    dat = dataset,
    cols = rainbow(100, end = 0.7),
    bg = "gray85"
  ) %>%
  add_osm_surface(
    roads,
    dat = dataset,
    cols = rainbow(100, end = 0.7) %>%
      adjust_colours(-0.2),
    bg = "gray85"
  ) %>%
  add_colourbar(
    cols = rainbow(100, end = 0.7),
    zlims = range(dataset$z)
  ) %>%
  add_axes()
```

## Pro and contra of `osmplotr` for continuous coloured maps

Note that `osmplotr` can create maps with discrete colours too. This was outside the scope of this blog post. See the `osmplotr` vignettes on plotting [basic maps](https://cran.r-project.org/web/packages/osmplotr/vignettes/basic-maps.html) and [data maps](https://cran.r-project.org/web/packages/osmplotr/vignettes/data-maps.html) (both categorical and continuous).

Much depends on the kind of data you want to display and the required accuracy. `osmplotr` uses the objects from OSM. Each OSM feature will get a uniform colour depending on the location of the centroid. Unless you data matched with the OSM data, it is not possible to use the accurate location of the data points.

Another point is scale and resolution of the OSM data. In case of an urban area were all individual buildings are available and the size of the individual buildings is small compared to the size of the map, then the `osmplotr` map works quite well.
