---
title: Estimating densities from a point pattern
author: Thierry Onkelinx
date: '2017-06-28'
slug: estimating-densities-from-a-point-pattern
categories: ["statistics", "point pattern"]
tags: ["point pattern", "density", "leaflet", "ggmap"]
---

In this example we focus on a set of 10450 coordinates in a small area. The goal is to estimate the local density of points, expressed as the number of point per unit area. The raw coordinates are given in [WGS84 (EPSG:4326)](https://epsg.io/4326), which is a geodetic coordinate system. That is not suited for calculating distances, so we need to reproject the points into a local projected coordinate system. In this case we use [Lambert72 (EPSG:3170)](https://epsg.io/31370). Next we calculate the density. To visualise the density, we have to transform the results back in to WGS84. 

The data used in this example is real data by centered to a different location for privacy reasons.

First we must read the data into R. Plotting the raw data helps to check errors in the data.

```{r read-data, fig.cap = "Raw data"}
points <- read.delim("points.txt", sep = " ")
library(leaflet)
leaflet(points) %>%
  addTiles() %>%
  addMarkers(lng = ~ lon, lat = ~lat)
```

The next step is to convert the dataset in to a `SpatialPoints` object with WGS84 project and reproject it into Lambert72. `sp::CRS()` defines the coordinate systems. `sp::coordinates()<-` is an easy way to convert a `data.frame` into a `SpatialPointsDataFrame`, but without specifying a coordinate system. Therefore we need to override the `proj4string` slot with the correct coordinate system. `sp::spTransform()` converts the spatial object from the current coordinate system to another coordinate system.

```{r reproject}
library(sp)
crs_wgs84 <- CRS("+init=epsg:4326")
crs_lambert <- CRS("+init=epsg:31370")
coordinates(points) <- ~lon + lat
points@proj4string <- crs_wgs84
points_lambert <- spTransform(points, crs_lambert)
```

