---
title: Nested and crossed random effects in `lme4`
author: Thierry Onkelinx
date: '2017-07-18'
slug: lme4-random-effects
categories: ["statistics", "mixed models"]
tags: ["lme4", "random effect"]
---

People often get confused on how to code nested and crossed random effects in the [`lme4`](https://cran.rstudio.com/web/packages/lme4/) package. I will try to make this more clear using some artificial datasets.

## Nested random effects

Nested random effects assume that there is some kind of hierarchy in the grouping of the observations. E.g. schools and classes. A class groups a number of students and a school groups a number of classes. There is a one-to-many relationship between the random effects. E.g. a school can contain multiple classes but a class can only be part of one school. Lets start by creating a simple example with fake data to explain the design.

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

```{r load-package}
library(DT)
library(lme4)
library(tidyverse)
library(htmlwidgets)
```

```{r nested-design}
set.seed(123)
n_school <- 10
mean_n_class <- 7
mean_n_student <- 5

n_class <- rpois(n_school, mean_n_class)
dataset <- map2_df(
  seq_len(n_school), 
  n_class, 
  ~data_frame(
    school = .x, 
    class = seq_len(.y),
    students = rpois(.y, mean_n_student)
  )
) %>%
  group_by(school, class) %>%
  do(
    student = data_frame(student = seq_len(.$students))
  ) %>%
  unnest(student) %>%
  mutate(
    class2 = interaction(school, class, drop = TRUE),
    student2 = interaction(class2, student, drop = TRUE)
  )

```

```{r echo = FALSE}
htmlwidgets::saveWidget(
  datatable(dataset, caption = "Design of the dataset"),
  file = "../../public/post/nested_crossed_nested_design.html", 
  selfcontained = TRUE
)
```

<iframe seamless src="../../../post/nested_crossed_nested_design.html" width="100%" height="500"></iframe>

`dataset` contains 3 design variables: `school`, `class` and `student`. Each numbering restarts at 1 when the higher level number changes. Hence the id of class and student are not unique. Therefore I added two new variables `class2` and `student2` which are unique id's for each class and student. The next step is adding the expected and observed values.

```{r nesting-values}
school_sd <- 2
class_sd <- 2
noise_sd <- 1
intercept <- 50

school_effect <- rnorm(n_school, mean = 0, sd = school_sd)
class_effect <- rnorm(length(levels(dataset$class2)), mean = 0, sd = class_sd)
dataset <- dataset %>%
  mutate(
    mu = intercept + school_effect[school] + class_effect[class2],
    y = mu + rnorm(n(), mean = 0, sd = noise_sd)
  )
```

### Explicit nesting

The first option is to use explicit nesting. Here we add a random effect for each hierarchical level and use the `:` notation to add all higher levels. This can be expanded to more than two levels. E.g. `(1|A) + (1|A:B) + (1|A:B:C) + (1|A:B:C:D)`. The nice thing about this notation is twofold: a) the nesting is explicit and clear for all readers; b) it is insensitive for the order: e.g. `(1|A) + (1|A:B)` is identical to `(1|B:A) + (1|A)`.

```{r}
lmer(y ~ (1|school) + (1|school:class), data = dataset)
lmer(y ~ (1|class:school) + (1|school), data = dataset)
```

### Shorthand nesting

`(1|A) + (1|A:B)` can be abbreviated into `(1|A/B)`. However, I recommend against it because here the order is important as seen in the example below. `(1|B/A)` expands to `(1|B) + (1|B:A)`, which is clearly a different model than `(1|A) + (1|A:B)`. I've seen many people being confused about the order, therefore I recommend to be explicit instead of using shorthand.

```{r}
lmer(y ~ (1|school/class), data = dataset)
lmer(y ~ (1|class/school), data = dataset)
```

### Implicit nesting

With implicit nesting, the nesting is 'defined' in the data. That is each level of a random effect has a one-to-many relation with the levels of the lower random effect. E.g. each class id is unique for a given class in a given school and cannot refer to a class in any other school. This is how we constructed the `class2` variable in our data. With implicit nesting the code can be abbreviated to `(1|A) + (1|B)`. Note that the `(1|A) + (1|A:B)` and `(1|A/B)` notations remain valid.

```{r}
lmer(y ~ (1|school) + (1|class2), data = dataset)
lmer(y ~ (1|school) + (1|school:class2), data = dataset)
lmer(y ~ (1|school/class2), data = dataset)
```

