---
title: A git workflow for ecologists
author: Thierry Onkelinx
date: '2017-08-23'
slug: git_workflow_ecologists
categories: ["reproducible research", "version control"]
tags: ["git"]
banner: ''
description: ''
images: []
menu: ''
output:
  blogdown::html_page:
    toc: TRUE
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  cache = TRUE,
  autodep = TRUE,
  echo = FALSE,
  warning = FALSE,
  fig.width = 6
)
library(dplyr)
library(diagram)
library(ggplot2)
```

# Git

For those how don't know [git](https://git-scm.com/), it is a free and open source distributed version control system designed to handle everything from small to very large projects with speed and efficiency. I use git daily, including for this [blog](https://github.com/thierryo/my_blog). Have a look at [Wikipedia](https://en.wikipedia.org/wiki/Git) for more background.

Although it requires some overhead, it save a lot of time once you get the hang of it. Why? Because you have the confidence that you can go back to any point in the history of a project. So you can explore new things without risking to ruin everything. The new things doesn't work out? Just go back to the last good point in the history and start over.

Each point in the history is called a `commmit`. A `commit` contains all essential information on _what_ needs to change to recreate the current state starting from the previous `commit`. It also contains useful metadata: _who_ created the `commit`, _when_ and _why^[Assuming that the user entered a sensible commit message.]_.

Git works great with plain text files like R scripts, RMarkdown files, data in txt or csv format, ... You can add binary files (Word, Excel, pdf, jpg, ...) to a git project, but not as efficient as plain text files and with less options. In case of a plain text file, git note which lines in the file are removed and where a given line was inserted. A change is a line is a combination of removing the old line and inserting the new line. Have a look a [this commit](https://github.com/ThierryO/my_blog/commit/fcab2804b75e7848283ef460f945f93aa3538bf5) if you want a real life example. Such granular approach is not available for binary files. Hence the old version is removed and the new version is added.

# Target audience for this workflow

The workflow is useful for anyone with basic computer skills. The workflow does not use all whisles and bells available in git. Only the minimal functionality which is all accessable via either a graphical user interface (GUI) or a website. We target ecologists who often write R scripts and have no prior knowledge on version control systems.

This workflow seems to work for a team of scientists how work on the same project and have all write access to that project (`repository` in git terminology).

# Basic workflow

## Use case

- First `repositories` of git novices.
- Initial start of a `repository`.

It is no longer valid as soon as more than one user commits to the `repository`.

## Principle

The basic workflow is just a simple linear history. The user makes a set of changes and commits those changes. This is repeated over and over until the project is finished. The resulting history will look like fig. \@ref(fig:basic).

One extra step is at least a daily `push` to another machine. This creates (or updates) a copy of the entire project history to that other machine. And thus serves as a backup copy. Therefore this should be done at least daily. The easiest way is to use an online service like [GitHub](https://github.com), [Bitbucket](https://bitbucket.com), [GitLab](https://gitlab.com), ... GitHub is free for public repositories. Bitbucket is offers free private repositories but only for small teams (max 5 users). GitHub is popular with freeware open source projects. Having the repository on an only platform has another benefit: it is easy to share your work with others or to work together with others.

```{r basic, fig.cap = "An example of the history of a basic workflow", fig.height = 1.5}
data.frame(
  x = 1:10,
  y = "master"
) %>%
  mutate(
    xend = lead(x) - 0.1,
    yend = y
  ) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_segment(
    arrow = arrow(length = unit(2, "mm"), type = "closed")
  ) +
  geom_point(aes(colour = y), size = 3, show.legend = FALSE) +
  theme(
    axis.text.x = element_blank(),
    axis.title.y = element_blank()
  ) +
  xlab("Time")
```

# Branching workflow with pull requests

## Use case

- Working with several people on the same repository
- More experienced git users

## Principle

1. Commits are only created in `feature branches`, not in the `master branch`.
1. Finalised `branches` are `merged` into the `master branch` by `pull requests`.

## Branch

The basic workflow has a single `branch` which is called `master`. Git makes it easy to create new `branches`. A `branch` starts from a specific commit. Each user should create a new `branch` when he starts working on a new feature in the repository. Because each user works in his own branch, he is the only one writing to this part of the history. This avoids a lot of conflicts. Fig. \@ref(fig:branching) illustrates how the history looks like when a few branches are created.

```{r branching, fig.cap = "An example of a history with a few feature branches", fig.height=3}
dataset <- frame_data(
  ~commit, ~parent, ~branch,
  1, NA, "master",
  2, 1, "master",
  3, 2, "feature A",
  4, 3, "feature A",
  5, 2, "feature B",
  6, 4, "feature A",
  7, 5, "feature B"
)
dataset %>%
  select(parent = commit, parent_branch = branch) %>%
  right_join(dataset, by = "parent") %>%
  ggplot(aes(x = commit, y = branch)) +
  geom_segment(
    aes(x = commit - 0.1, xend = parent, yend = parent_branch),
    arrow = arrow(length = unit(2, "mm"), type = "closed", ends = "first")
  ) +
  geom_point(aes(colour = branch), size = 3, show.legend = FALSE) +
  theme(
    axis.text.x = element_blank(),
    axis.title.y = element_blank()
  ) +
  xlab("Time")
```

## Pull request

Creating branches is fine, but they diverge the history of the repository. So we need a mechanism to `merge` branches together. In this workflow we will work on a feature branch until it is finished. Then we merge it into the master branch. Fig. \@ref(fig:pull-request) illustrates the resulting history. This can be done locally using a `merge`, but it is safer to do it online via a `pull request`.

```{r pull-request, fig.cap = "An example of a history after two pull requests", fig.height=3}
dataset <- frame_data(
  ~commit, ~parent, ~branch,
  1, NA, "master",
  2, 1, "master",
  3, 2, "feature A",
  4, 3, "feature A",
  5, 2, "feature B",
  6, 4, "feature A",
  7, 5, "feature B",
  8, 2, "master",
  8, 6, "master",
  9, 8, "feature B",
  9, 7, "feature B",
  10, 8, "master",
  10, 9, "master",
  11, 10, "feature C"
)
dataset %>%
  select(parent = commit, parent_branch = branch) %>%
  right_join(dataset, by = "parent") %>%
  ggplot(aes(x = commit, y = branch)) +
  geom_segment(
    aes(x = commit - 0.1, xend = parent, yend = parent_branch),
    arrow = arrow(length = unit(2, "mm"), type = "closed", ends = "first")
  ) +
  geom_point(aes(colour = branch), size = 3, show.legend = FALSE) +
  theme(
    axis.text.x = element_blank(),
    axis.title.y = element_blank()
  ) +
  xlab("Time")
```

A `pull request` is a two step procedure. First you create the `pull request` by indication online which branches you would like to `merge`. The second step is to `merge` the pull request. Documentation on how to handle `pull requests` can be found on the websites of [GitHub](https://help.github.com/articles/about-pull-requests/), [Bitbucket](https://www.atlassian.com/git/tutorials/making-a-pull-request) and [GitLab](https://docs.gitlab.com/ee/gitlab-basics/add-merge-request.html).

Pull requests have several advantages over local merges

1. It works only when the branches are pushed to the online copy of the repository. This ensures not only a backup but also gives access to the latest version to your colaborators.
1. All pull requests are done against the common (online) master branch. Local merges would create diverging master branches which will create a lot of conflicts.
1. Since the pull request is a two step procedure, one user can create the pull request and another (e.g. the project leader) can do the actual merge.
1. The pull request gives an overview of the aggregate changes of all the commits in the pull request. This makes it easier to get a feeling on what has been changed within the range of the pull request.
1. Most online tools allow to add comments and reviews to a pull request. This is useful to discuss a feature prior to merging it. In case some changes are required, then the user should update his feature branch. The pull request get automatically updated.

## Conflicts

Conflicts arise when a file is changed at the same location in two different branches and with different changes. Git cannot descide which version is correct and therefore blocks the merging of the pull request. It is up to the user to select the correct version and commit the required changes. See online [tutorials](https://help.github.com/articles/resolving-a-merge-conflict-using-the-command-line/) on how to do this. Once the conflicts are resolved, you can go ahead and merge the pull request. This is illustrated in fig. \@ref(fig:pull-request). First `master` is merged back into `feature B` to handle the merge conflict and then `feature B` is merged into `master`.

_What if I choose the wrong version?_ Don't panic, both versions remain in the history so you don't loose any. So you can create a new branch starting for the latest commit with the correct version and merge that branch.

# Flowcharts

Here a a fex flowcharts that illustrate several components of the branching workflow with pull requests. Fig. \@ref(fig:prepare-repo) illustrates the steps you need when you want to start working on a project. Once you have a local `clone` of the repository you can `check out` the required feature branch (fig. \@ref(fig:create-branch)).

```{r prepare-repo, fig.cap = "Flowchart for preparing a repository."}
par(mar = rep(0, 4))
openplotmat()

straightarrow(c(0.1, 0.9), c(0.5, 0.9))
straightarrow(c(0.5, 0.9), c(0.85, 0.9))
straightarrow(c(0.5, 0.9), c(0.5, 0.5))
straightarrow(c(0.5, 0.5), c(0.1, 0.5))
straightarrow(c(0.5, 0.5), c(0.5, 0.15))

bentarrow(c(0.9, 0.85), c(0.5, 0.5), path = "V")
bentarrow(c(0.1, 0.5), c(0.5, 0.15), path = "V")

text(0.70, 0.95, "no")
text(0.55, 0.7, "yes")
text(0.30, 0.55, "no")
text(0.55, 0.3, "yes")

textellipse(c(0.1, 0.9), radx = 0.1, lab = "Start")
textdiamond(c(0.5, 0.9), radx = 0.1, rady = 0.1, lab = "Repository\nexists?")
textrect(c(0.85, 0.9), radx = 0.1, lab = "Create\nrepository\nonline")
textdiamond(c(0.5, 0.5), radx = 0.1, rady = 0.1, lab = "Local clone\nexists?")
textrect(c(0.1, 0.5), radx = 0.1, lab = "Clone\nrepository")
textellipse(c(0.5, 0.15), radx = 0.1, lab = "Check out\nfeature branch")
```

```{r create-branch, fig.cap = "Flowchart for changing to a feature branch."}
par(mar = rep(0, 4))
openplotmat()

straightarrow(c(0.1, 0.9), c(0.5, 0.9))
straightarrow(c(0.5, 0.9), c(0.85, 0.9))
straightarrow(c(0.5, 0.9), c(0.5, 0.5))
straightarrow(c(0.5, 0.5), c(0.1, 0.5))
straightarrow(c(0.85, 0.9), c(0.85, 0.5))
straightarrow(c(0.1, 0.5), c(0.1, 0.15))
straightarrow(c(0.85, 0.5), c(0.85, 0.15))
straightarrow(c(0.85, 0.15), c(0.5, 0.15))
straightarrow(c(0.5, 0.15), c(0.1, 0.15))

text(0.70, 0.95, "no")
text(0.55, 0.7, "yes")

textellipse(c(0.1, 0.9), radx = 0.1, lab = "Open\nrepository")
textdiamond(c(0.5, 0.9), radx = 0.1, rady = 0.1, lab = "Feature\nbranch\nexists?")
textrect(c(0.85, 0.9), radx = 0.1, lab = "Check out\nmaster branch")
textrect(c(0.5, 0.5), radx = 0.1, lab = "Check out\nfeature branch")
textrect(c(0.1, 0.5), radx = 0.1, lab = "Pull")
textrect(c(0.85, 0.5), radx = 0.1, lab = "Pull")
textrect(c(0.85, 0.15), radx = 0.1, lab = "Create and\ncheck out\nfeature branch")
textrect(c(0.5, 0.15), radx = 0.1, lab = "Push")
textellipse(c(0.1, 0.15), radx = 0.1, lab = "Start\nworking")
```

